Modelling writing hesitations in text writing as finite mixture process

### 2023-07-17 report/manuscript.Rmd
Add manuscript and simulation

---
title: "Modelling writing hesitations in text writing as finite mixture process"
author: "Jens Roeser"
date: "Compiled `r format(Sys.Date(), '%b %d %Y')`"
output: 
  rmdformats::downcute:
    keep_md: true
    self_contained: true
    thumbnails: false # for images
    lightbox: true
    gallery: false
    highlight: tango
    use_bookdown: true # for cross references
bibliography: references.bib
csl: apa.csl
link-citations: yes
---

```{r include = F}
library(tidyverse)
library(kableExtra)
library(janitor)
library(brms)
library(polspline)
library(patchwork)
library(rmdformats)
library(gridExtra)
library(grid)
library(gtable)
library(scales)
library(ggthemes)
source("../scripts/functions.R")
source("scripts/get_pdens_plot.R")

theme_set(theme_bw(base_size = 12) +
            theme(strip.background = element_blank(),
                  legend.position = "top",
                  legend.justification = "right",
                  panel.grid = element_blank(),
                  panel.background = element_rect(fill = "transparent"), # bg of the panel
                  plot.background = element_rect(fill = "transparent", color = NA)))

@@

beta_label <- "Short key transition\nduration"
delta_label <- "Slowdown for hesitant key\ntransitions"
theta_label <- "Probability of hesitant\ntransitions"
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      comment = NA,
                      #                      dev = c("pdf", "png"), 
                      #                     fig.keep = "all", 
                      #                    fig.path = "figures/",
                      fig.align = 'center', 
                      #                   pdf.options(encoding = "ISOLatin9.enc"),
                      fig.width = 8, 
                      fig.height = 5,
                      width=90)
```

@@

```{r eval = F}
Take intro from Mark's draft of SPL2 paper

Different views on pausing in text production, largely representing typing as a unimodal or bimodal process

- pauses are represented in the right tail of the distribution (unimodal) explaining the long positive skew and therefore possibly making a log transformation redundant (there are people arguing against log transformation of keystrokes but I don't understand their argument)
- as pauses are more likely at linguistic edges, the variance of the distribution is going to be larger
- pauses are represented as an independent distribution / process
  - short distribution: fluent key transitions only limited by the ability to move fingers (same across text locations)
  - short distribution: processes involved in cascade are different when lexical and syntactic planning is involved (differences across text locations)

# Model definitions

@@

## Uni-modal Gaussian

@@

$$
\begin{align}
\text{iki}_i & \sim \text{N}(\mu_i, \sigma_\text{e}^2)\\
\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
\end{align}
$$

## Uni-modal log-Gaussian

@@

$$
\begin{align}
\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_\text{e}^2) \\
\text{where: } &
\mu_i = \beta_\text{location[i]} + \text{participant}_i\\
& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
\end{align}
$$

## Uni-modal unequal-variance log-Gaussian

$$
\begin{align}
\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
\end{align}
$$

## Bi-modal log-Gaussian (constrained)

@@

$$
\begin{align}
\text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
\text{where: }
		&\delta \sim \text{N}(0,1)\\
 	 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta > 0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{align}
$$
This models takes into account two source = s of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.

## Bi-modal log-Gaussian (unconstrained)

@@

$$
\begin{align}
\text{iki}_{i} &\sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
	\text{where: } & \delta \sim \text{N}(0,1)\\
	 	 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta > 0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{align}
$$

# Analysis

We reanalysed data sets including process information from participants writing text. For all data sets we fit a series of four models each with random effects for participants. Probability functions used were normal and log-normal in line with typically treatments used in the literature, a log-normal distribution with unequal variances for model predictions, and a bimodal mixed effects model. Stan code for mixture models was based on @roeser2021modelling. Text locations (levels: before sentence, before word, within word) was included as predictor in all models.

# Datasets

## General overview

@@

```{r datasets, results='asis'}
tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT"),
                Source = c("@ronneberg2022process",
                              "@torrance2016adolescent",
                              "@torrance",
                              "@rossetti2022s",
                              "@nina_vandermeulen_2020_3893538"),
                Keylogger = c("EyeWrite", 
                           "EyeWrite", 
                           "CyWrite", 
                           "InputLog", 
                           "InputLog"),
                `Writing task` = c(
                                 "Argumentative essays", 
                                 "Expository texts",
                                 "Argumentative essays",
                                 "Text simplification",
                                 "Synthesis"),
                 `N (ppts)` = c(126, 26*2, 39, 47, 658),
#                n_texts = c(1, 2, 2, 2, NA),
#                n_sentences = c(),
#                n_words = c(),
                conditions = c(" ", 
                               "weak decoders / control; masked / unmasked",
                               "write in L1 / L2",
                               "pre / post test trained in plain language principles and control", 
                               "Various topics and genres"),
                `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95),
                 Language = c( "Norwegian", 
                               "Norwegian", 
                               "English (L1) / Spanish (L2)", 
                               "English (L2)", 
                               "Dutch")) %>% 
  kable(caption = "Datasets in brief.") %>%
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 12)

@@

##  Transition types

@@

## Data reduction

```{r reductionfunctions}
# Function to get data reduction infos
get_lift <- function(file, n_samples, n_ppts){
#  file <- "../data/lift.csv"
# n_samples = 50
#  n_ppts = 100
  result <- list()

  d <- read_csv(file)

  full_set <- d %>% 
    count(ppt, topic, genre) %>% 
    summarise(n = n(), .by = ppt) %>% 
    mutate(has_full_set = n == 4)

  result[[3]] <- full_set %>% count(has_full_set)

  ppt_keep <- full_set %>% 
    filter(has_full_set) %>% 
    pull(ppt)

  result[[4]] <- d %>% 
    filter(ppt %in% ppt_keep) %>% 
    select(ppt, enough_sentences) %>% 
    unique() %>% 
    count(enough_sentences)

  result[[1]] <- d %>%
    filter(!is.na(iki), 
           !is_edit,
           ppt %in% ppt_keep,
           enough_sentences) %>%
    select(ppt, iki, location) %>% 
    drop_na() %>% 
    mutate(too_fast = iki <= 50, 
           too_slow = iki >= 30000,
           across(ppt, ~as.numeric(factor(.)))) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean, 
                          se = se_bin)),
              .by = c(ppt)) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean), 
                     .names = "{.col}"))

  # Sample within each category random data points
  set.seed(365)
  result[[2]] <- d %>% 
    filter(!is.na(iki), 
           !is_edit,
           ppt %in% ppt_keep,
           enough_sentences,
           iki > 50,
           iki < 30000) %>% 
    mutate(keep = 1:n(),
           keep = sample(keep),
           keep = keep <= n_samples,
           .by = c(ppt, location, genre, topic)) %>% 
   summarise(across(keep, 
             list(mean = mean, 
                  se = se_bin)),
              .by = c(ppt, location)) %>% 
   summarise(across(starts_with("keep"), 
                     list(mean = mean), 
                     .names = "{.col}"), 
              .by = location)

   set.seed(365)
   result[[5]] <- d %>% 
    mutate(keep = 1:n(),
           keep = sample(keep),
           keep = keep <= n_samples, 
           .by = c(ppt, location, genre, topic)) %>% 
    filter(keep) %>% 
    mutate(ppt_keep = ppt %in% sample(unique(ppt), n_ppts)) %>% 
    select(ppt, ppt_keep) %>% 
    unique() %>% 
    count(ppt_keep)

   return(result)
}
get_c2l1 <- function(file, n_samples){
  #file <- "../data/c2l1.csv"
  #n_samples <- 100
  d <- read_csv(file) %>% 
    filter(!is.na(iki)) %>% 
    mutate(ppt = as.numeric(factor(subno))) %>% 
    select(ppt, iki, location)

  result <- list()

  result[[1]] <- d %>% 
    mutate(too_fast = iki <= 50,
           too_slow = iki >= 30000) %>%
    summarise(across(starts_with("too"), 
                     list(mean = mean, se = se_bin)),
              .by = c(ppt)) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean), 
                     .names = "{.col}"))

  set.seed(365)
  result[[2]] <- d %>% 
    filter(iki > 50,
           iki < 30000) %>% 
    mutate(keep = 1:n(),
           keep = sample(keep),
           keep = keep <= n_samples,
           .by = c(ppt, location)) %>% 
   summarise(across(keep, 
             list(mean = mean, se = se_bin)),
              .by = c(ppt, location)) %>% 
    summarise(across(starts_with("keep"), 
                     list(mean = mean), 
                     .names = "{.col}"), 
              .by = location)

  return(result)
}
get_cato <- function(file, n_samples){
  d <- read_csv(file) %>% 
    filter(!is.na(iki)) %>% 
    mutate(ppt = as.numeric(factor(subno))) %>% 
    select(ppt, iki, location, xn, dystyp)

  result <- list()

  result[[1]] <- d %>% 
    mutate(too_fast = iki <= 50,
           too_slow = iki >= 30000) %>%
    summarise(across(starts_with("too"), 
                     list(mean = mean, 
                          se = se_bin)),
              .by = c(ppt)) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean), 
                     .names = "{.col}"))

  set.seed(365)
  result[[2]] <- d %>% 
    filter(iki > 50,
           iki < 30000) %>% 
    mutate(keep = 1:n(),
           keep = sample(keep),
           keep = keep <= n_samples,
           .by = c(ppt, location, xn, dystyp)) %>% 
   summarise(across(keep, 
             list(mean = mean, 
                  se = se_bin)),
              .by = c(ppt, location)) %>% 
   summarise(across(starts_with("keep"), 
                     list(mean = mean), 
                     .names = "{.col}"), 
              .by = location)

  return(result)
}
get_plantra <- function(file, n_samples){
  d <- read_csv(file) %>% 
    filter(!is.na(iki), !is_edit) %>% 
    mutate(ppt = as.numeric(factor(ppt))) %>% 
    select(ppt, iki, location, task)

  result <- list()

  result[[1]] <- d %>% 
    mutate(too_fast = iki <= 50,
           too_slow = iki >= 30000) %>%
    summarise(across(starts_with("too"), 
                     list(mean = mean, 
                          se = se_bin)),
              .by = c(ppt, task)) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean), 
                     .names = "{.col}"))

  # keep only participants that did both the pre and post task.
  keep_ppts <- select(d, ppt, task) %>% 
    unique() %>% 
    count(ppt) %>% 
    mutate(both_tasks = n == 2)

  result[[3]] <- count(keep_ppts, both_tasks)

  keep_ppts <- keep_ppts %>% filter(both_tasks) %>% pull(ppt)
  d <- filter(d, ppt %in% keep_ppts)

  # Filter ppts with too few sentences
  keep_ppts <- count(d, ppt, location) %>% 
    filter(location == "before sentence") %>% 
    mutate(enough_sentences = n >= 10)

  result[[4]] <- count(keep_ppts, enough_sentences) %>% 
    rename(more_than_10_sentences = enough_sentences)

  keep_ppts <- keep_ppts %>% 
    filter(enough_sentences) %>% 
    pull(ppt)

  d <- filter(d, ppt %in% keep_ppts)

  set.seed(365)
  result[[2]] <- d %>% 
    filter(iki > 50,
           iki < 30000) %>% 
    mutate(keep = 1:n(),
           keep = sample(keep),
           keep = keep <= n_samples,
           .by = c(ppt, location, task)) %>% 
   summarise(across(keep, 
             list(mean = mean, 
                  se = se_bin)),
              .by = c(ppt, location)) %>% 
   summarise(across(starts_with("keep"), 
                     list(mean = mean), 
                     .names = "{.col}"), 
              .by = location)
  return(result)
}
get_spl2 <- function(file, n_samples){
  result <- list()
#  file <- "../data/spl2.csv"
  d <- read_csv(file) %>% 
    select(-transition_dur) %>% 
    rename(iki = transition_dur_to_mod,
           location = transition_type) %>% 
    filter(!is.na(iki), 
           edit == "noedit") %>% 
    select(ppt = SubNo, iki, location, Lang) %>% 
    drop_na()

  result[[1]] <- d %>% 
    mutate(too_short = iki <= 50,
           too_fast = iki >= 30000,
           across(ppt, ~as.numeric(factor(.)))) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean, 
                          se = se_bin)),
              .by = c(ppt)) %>% 
    summarise(across(starts_with("too"), 
                     list(mean = mean), 
                     .names = "{.col}"))

  d <- d %>% filter(iki > 50, iki < 30000)

  # Filter ppts with too few sentences
  keep_ppts <- count(d, ppt, location) %>% 
    filter(location == "sentence_before") %>% 
    mutate(enough_sentences = n >= 10)

  result[[3]] <- count(keep_ppts, enough_sentences) %>% 
    rename(more_than_10_sentences = enough_sentences)

  keep_ppts <- keep_ppts %>% 
    filter(enough_sentences) %>% 
    pull(ppt)

  d <- filter(d, ppt %in% keep_ppts)

  set.seed(365)
  result[[2]] <- d %>% 
    mutate(keep = 1:n(),
           keep = sample(keep),
           keep = keep <= n_samples,
           .by = c(ppt, location, Lang)) %>% 
   summarise(across(keep, 
             list(mean = mean, 
                  se = se_bin)),
              .by = c(ppt, location)) %>% 
   summarise(across(starts_with("keep"), 
                     list(mean = mean), 
                     .names = "{.col}"), 
              .by = location)

  return(result)
}
```

```{r}
# Get info about data reduction
n_samples <- 100
c2l1 <- get_c2l1("../data/c2l1.csv", n_samples = n_samples)
cato <- get_cato("../data/cato.csv", n_samples = n_samples)
lift <- get_lift("../data/lift.csv", n_samples = 50, n_ppts = 100)
plantra <- get_plantra("../data/plantra.csv", n_samples = n_samples)
spl2 <- get_spl2("../data/spl2.csv", n_samples = n_samples)

@@

```{r eval = F}
Data reduction using random samples of subsets of participants for larger data sets

Also for larger data sets, only used participants that provided a min number of sentences

Removed transitions that were followed by an editing operation

Extremely short and long intervals

Add a description with general data filtering and a table with how many data were removed from the analysis by study.

Add final sample number of ppt, and transitions by location

# Out-of-samples cross-validation

@@

# Cross-data set comparisons

@@

## Cross-data set visualisation

@@

## Effect of transition location

It's generally believed that pausing is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges, i.e. before sentence > before word > within word. We have evaluated the differences between transition locations for all data sets. The results are shown in Table \@ref(tab:loceffect).

@@

```{r eval = F}

#
# Do I really need all this?
#
#
#
#

# use posterior to calculate interactions, cell means and dataset differences
files <- list.files(str_c("../stanout/", 
                          c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1")), 
                    pattern = "mogbetaun.+.csv", 
                    full.names = T)

plantra <- read_csv(files[1], id = "data") %>% 
  pivot_wider(names_from = c(genre, topic), values_from = value) %>% 
  unnest(cols = c(ARG_A, ARG_B, ARG_C, ARG_D, INF_A, INF_B, INF_C, INF_D)) %>% 
  rowwise() %>% 
  mutate(value = mean(c_across(contains("_")))) %>% 
  ungroup() %>% 
  select(param, location, value, data)

spl2 <- read_csv(files[2], id = "data") %>% 
  pivot_wider(names_from = c(lang), values_from = value) %>% 
  unnest(cols = c(EN, ES)) %>% 
  rowwise() %>% 
  mutate(value = mean(c_across(c(EN, ES)))) %>% 
  ungroup() %>% 
  select(param, location, value, data)

lift <- read_csv(files[3], id = "data") %>% 
  pivot_wider(names_from = c(task), values_from = value) %>% 
  unnest(cols = c(`pre-test`, `post-test`)) %>% 
  rowwise() %>% 
  mutate(value = mean(c_across(contains("-test")))) %>% 
  ungroup() %>% 
  select(param, location, value, data)

cato <- read_csv(files[4], id = "data") %>% 
  pivot_wider(names_from = c(task, group), values_from = value) %>% 
  unnest() %>% 
  rowwise() %>% 
  mutate(value = mean(c_across(contains("_")))) %>% 
  ungroup() %>% 
  select(param, location, value, data)

c2l1 <- read_csv(files[5], id = "data") %>% 
  select(param, location, value, data)

ps <- bind_rows(plantra, spl2, lift, cato, c2l1) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta, delta, prob)) %>% 
  mutate(across(data, 
                str_remove_all, 
                "\\.|\\/|stanout|\\mog.csv"),
         theta = logit_scaled(prob)) %>% 
  select(-prob) %>%
  pivot_longer(beta:theta) %>%
  pivot_wider(names_from = c(data, location), 
              values_from = value, 
              names_sep = "_") %>% 
  unnest(contains("_"))

>>>>

---
title             : "Modelling writing hesitations in text writing as finite mixture process"
shorttitle        : "Modelling writing hesitations"

csl               : "apa.csl"

author: 
  - name          : "Jens Roeser"
    affiliation   : "1"
    address       : "50 Shakespeare St, Nottingham NG1 4FQ"
    corresponding : yes 
    email         : "jens.roeser@ntu.ac.uk"
  - name          : "Mark Torrance"
    affiliation   : "1"
  - name          : "Rianne Conijn"
    affiliation   : "2"
  - name          : "Evgeny Chukharev-Hudilainen"
    affiliation   : "3"

affiliation:
  - id            : "1"
    institution   : "Department of Psychology, Nottingham Trent University, United Kingdom"

  - id            : "2"
    institution   : "Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands"

  - id            : "3"
    institution   : "Department of English, Iowa State University, Iowa"

abstract: |

keywords: "Keystroke modelling; finite mixture models; Bayesian models; text composition"

bibliography      : ["references.bib"]

documentclass     : "apa7"
classoption       : "man"
output            : 
  papaja::apa6_pdf:
    keep_tex: TRUE

figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
csquotes          : true

header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{graphicx}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{threeparttable}
  - \usepackage[normalem]{ulem}
  - \usepackage[utf8]{inputenc}
  - \usepackage{icomma}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
---

```{r packages, include=FALSE}
library(tidyverse)
library(kableExtra)
library(janitor)
library(brms)
library(polspline)
library(patchwork)
library(rmdformats)
library(gridExtra)
library(grid)
library(gtable)
library(scales)
library(ggthemes)
source("../scripts/functions.R")
source("scripts/get_pdens_plot.R")

theme_set(theme_few(base_size = 9) +
            theme(strip.background = element_blank(),
                  legend.position = "top",
                  legend.justification = "right",
                  panel.grid = element_blank(),
                  panel.background = element_rect(fill = "transparent"), # bg of the panel
                  plot.background = element_rect(fill = "transparent", color = NA)))

@@

beta_label <- "Short key transition\nduration"
delta_label <- "Slowdown for hesitant key\ntransitions"
theta_label <- "Probability of hesitant\ntransitions"

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      comment = NA,
                      cache.extra = R.version,
                      dev = "cairo_pdf",
                      fig.align = 'center', 
                      fig.width = 8, 
                      fig.height = 5,
                      width=90)

options(kableExtra.auto_format = FALSE)
dev.args <- list(pdf = list(type = "cairo"))
```

@@

# Methods

## Statistical models

@@

### Uni-modal Gaussian

@@

$$
\begin{align}(\#eq:unimodgaus)
\text{iki}_i & \sim \text{N}(\mu_i, \sigma_\text{e}^2)\\
\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
\end{align}
$$

### Uni-modal log-Gaussian

@@

$$
\begin{align}(\#eq:unimodloggaus)
\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_\text{e}^2) \\
\text{where: } &
\mu_i = \beta_\text{location[i]} + \text{participant}_i\\
& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
\end{align}
$$

### Uni-modal unequal-variance log-Gaussian

$$
\begin{align}(\#eq:unimoduv)
\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
\end{align}
$$

### Bi-modal log-Gaussian (constrained)

@@

$$
\begin{align}(\#eq:bimodcon)
\text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
\text{where: }
		&\delta \sim \text{N}(0,1)\\
 	 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta > 0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{align}
$$
This models takes into account two source = s of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.

### Bi-modal log-Gaussian (unconstrained)

@@

$$
\begin{align}(\#eq:bimoduncon)
\text{iki}_{i} &\sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
	\text{where: } & \delta \sim \text{N}(0,1)\\
	 	 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta > 0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{align}
$$

## Data sets

@@

```{r datasets, results='asis'}
tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT", "GUNNEXP2"),
                Source = c("@ronneberg2022process",
                              "@torrance2016adolescent",
                              "@torrance",
                              "@rossetti2022s",
                              "@nina_vandermeulen_2020_3893538", 
                           ""),
                Keylogger = c("EyeWrite", 
                           "EyeWrite", 
                           "CyWrite", 
                           "InputLog", 
                           "InputLog",
                           "EyeWrite"),
                `Writing task` = c(
                                 "Argumentative essays", 
                                 "Expository texts",
                                 "Argumentative essays",
                                 "Text simplification",
                                 "Synthesis", 
                                 ""),
                 `N (ppts)` = c(126, 26*2, 39, 47, 658, 45),
#                n_texts = c(1, 2, 2, 2, NA),
#                n_sentences = c(),
#                n_words = c(),
                conditions = c(" ", 
                               "weak decoders / control; masked / unmasked",
                               "write in L1 / L2",
                               "pre / post test trained in plain language principles and control", 
                               "Various topics and genres",
                               "masked / unmasked"),
                `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95, NA),
                 Language = c( "Norwegian", 
                               "Norwegian", 
                               "English (L1) / Spanish (L2)", 
                               "English (L2)", 
                               "Dutch",
                               "Norwegian")) %>% 
  kable(caption = "Datasets in brief.") %>%
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 12)

@@

### GUNNEXP2

### C2L1

The C2L1 data set comprises data Norwegian 6th graders -- *N*=126, mean age 11 years 10 months -- published in @ronneberg2022process. The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.

### CATO

### PLanTra

### LIFT

### SPL2

## Data analysis

### Transition types

@@

### Data reduction

```{r reductionfunctions}
source("scripts/get_data_summary.R")
```

```{r}
# Get info about data reduction
n_samples <- 100
c2l1 <- get_c2l1("../data/c2l1.csv", n_samples = n_samples)
cato <- get_cato("../data/cato.csv", n_samples = n_samples)
lift <- get_lift("../data/lift.csv", n_samples = 50, n_ppts = 100)
plantra <- get_plantra("../data/plantra.csv", n_samples = n_samples)
spl2 <- get_spl2("../data/spl2.csv", n_samples = n_samples) 
gunnexp2 <- get_gunnexp2("../data/gunnexp2.csv", n_samples = n_samples)
```

For all datasets we only used transitions that were not followed by an editing operation.

We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set). We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant)

We removed keystroke intervals that are extremely short ($\le$ 50 msecs) or extremely long ($\ge$ 30 secs). The percentage of remove keystroke data can be found in Table \@ref(tab:datareduction).

From the remaining data we randomly sampled `r n_samples` observations per participant, per condition, and per transition location, with the exception of the LIFT data set. This was done for computational reasons to reduce the time the Bayesian models need to complete. For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our analysis. Because we included the large number of writing tasks in the LIFT data set as fixed effect, we sampled 50 observations per condition, location and participant. The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \@ref(tab:datareduction).

```{r datareduction, results='asis'}
c2l1$ds <- "c2l1"
cato$ds <- "cato"
lift$ds <- "lift"
plantra$ds <- "plantra"
spl2$ds <- "spl2"
gunnexp2$ds <- "gunnexp2"

extrem_values <- map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[1]] %>% 
          mutate(ds = .x$ds)) %>% 
  mutate(across(where(is.numeric), ~round(.*100, 2)),
         across(starts_with("too"), ~str_c(., "%"))) %>% 
  unite("<50ms", starts_with("too_fast"), sep = " (") %>% 
  unite(">30,000ms", starts_with("too_slow"), sep = " (") %>% 
  mutate(across(c(`<50ms`, `>30,000ms`), ~str_c(., ")"))) %>% 
  rename(`$\\le$ 50 msecs` = `<50ms`,
         `$\\ge$ 30 secs` = `>30,000ms`)

random_sample <- 
  map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[2]] %>% 
          mutate(ds = .x$ds)) %>% 
    mutate(across(where(is.numeric), ~round(.*100, 1)),
           across(starts_with("keep"), ~str_c(., "%"))) %>%
    unite("keep", starts_with("keep"), sep = " (") %>% 
    mutate(across(c(keep), ~str_c(., ")"))) %>%
  pivot_wider(names_from = location, values_from = keep)

datareduction <- left_join(extrem_values, random_sample, by = "ds") %>% 
  relocate(ds) %>% 
  mutate(across(ds, recode, 
                "cato" = "CATO",
                "spl2" = "SPL2", 
                "spl2_shift" = "SPL2 (shift + C)", 
                "plantra" = "PLanTra",
                "lift" = "LIFT",
                "c2l1" = "C2L1",
                "gunnexp2" = "GUNNEXP2")) %>% 
  rename(`Dataset` = ds)

kable(datareduction, 
      caption = "Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition locattion. Standard error is shown in parentheses.",
        align =c("l", rep("r", 14))) %>%
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 11) %>%
  add_header_above(c(" " = 1, 
                     "Extreme values" = 2,
                     "Randomly sampled data" = 3)) %>%
  column_spec(1, width = "8em")

@@

### Bayesian modelling

Keystroke data from text composition tasks were modelled in a series of five models. An overview of all models can be found in Table \ref{tab:models}.

```{r models, results = 'asis'}
models <- tibble(Models = paste0("M",1:4),
       Type = c("LMM", "LMM", "LMM", "MoG", "MoG"),
       Equation = paste0("\ \\ref{eq:", c("unimodgaus", "unimodloggaus", "unimoduv", "bimodcon", "bimoduncon"), "}"),
       Description = c("Gaussian distribution",
                       "Log-normal distribution",
                       "Log-normal distribution and unequal variances",
                       "Fluent transition are not allowed to vary by transition location",
                       "Fluent transition are allowed to vary by transition location"))

papaja::apa_table(models, align = c("l", "l", "l"), escape = FALSE, digits = 0, placement = "bp!",
                  caption = "Overview of typing-process models. All models were fitted with random intercepts for participants.",
                  note = "LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians") 
```

each with random effects for participants. Probability functions used were normal and log-normal in line with typically treatments used in the literature, a log-normal distribution with unequal variances for model predictions, and a bimodal mixed effects model. Stan code for mixture models was based on @roeser2021modelling. Text locations (levels: before sentence, before word, within word) was included as predictor in all models.

# Results

## Out-of-samples cross-validation

@@

## Cross-data set comparisons

@@

@@

It's generally agreed that pausing is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges, i.e. before sentence > before word > within word. We have evaluated the differences between transition locations for all data sets. The results are shown in Table \@ref(tab:loceffect).

@@

## Simulation

A general concern with mixture models is that in principle, as the mixture model has more parameters it might simply always lead to a better fit, even though cross-validation is addressing potential problems with overfitting models.

To address this concern we simulated two sets of data. Both data sets have two conditions and 1,000 observations each. The first set of data has as underlying data generating process a mixture model with two mixture components similar to the process described above. The difference between the two conditions is that the mixing proportion is larger for condition 2 than for condition 1, hence long observations are more likely in condition 2. This model can be summarised as followed:

$$
\begin{align}(\#eq:simmog)
\text{y}_i \sim\text{ } & \theta_\text{condition[i]} \cdot \text{logN}(\mu_1, \sigma^2_1) +\\
& (1 - \theta_\text{condition[i]}) \cdot \text{logN}(\mu_2, \sigma^2_2)
\end{align}
$$

The second data set was generated with an unequal variance unimodal model as data generating process. Condition 2 has a larger mean and standard deviation than condition 1. The model can be summaried as followed:

$$
\begin{align}(\#eq:simuv)
\text{y}_i \sim\text{ }& \text{logN}(\mu_\text{condition[i]}, \sigma^2_\text{condition[i]})
\end{align}
$$

The true parameter values used for each of the two data simulations can be found in Table \@ref(tab:simparam). The simulated data are visualised in Figure \@ref(fig:simdata). The data are simulated to be similarly distributed to keystroke transitions.

```{r simdata, fig.cap= "Data simulated with a bimodal process (left) and a unimodal process (right).", fig.height=4}
files <- list.files("../mixture-model-sim/data", full.names = T)

map_dfr(files, ~read_csv(.) %>% 
          mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
  mutate(across(condition, factor),
         across(condition, ~str_c("Condition: ", condition)),
         across(Dataset, ~recode(., mogdata = "bimodal", uvlmdata = "unimodal"))) %>% 
  ggplot(aes(x = value, colour = condition, fill = condition)) +
  geom_density(alpha = .25) +
  facet_grid(~Dataset, labeller = label_both) +
  scale_colour_colorblind() + 
  scale_fill_colorblind() +
  labs(colour = "", fill = "", x = "Simulated data", y = "Density")
